Region-Network Hierarchical Sparsity Priors for\\ High-Dimensional Inference in Brain Imaging

### 2016-02-04 paper/icml2016/example_paper.tex
email
sample complexity: add more results

\paragraph{Hyperparameter optimization.}
{\color{red} NEEDS EMBELLISHING}
We perform nested cross-validation to obtain optimal hyperparameters. 
In order to analyse their effect in detail, we split our analysis into two 
parts.
\begin{enumerate}
\item In the first part, we optimize the trade-off between the different hierarchy levels as well as the global penalty level for maximal prediction.
\item in the second part, we select the optimal global penalty level for 
given trade-off ratios.
\end{enumerate}

\paragraph{Control methods}
{\color{red} needs embellishing}
We evaluate our tree structure against other commonly used structured sparse
penalties.

@@

\paragraph{Benchmarking hierarchical tree sparsity against
common sparsity penalties.}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation.
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with sparse group sparsity
imposes a structured $\ell_1/\ell_2$ block norm with additional $\ell_1$ term
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
The stratified and shuffled training data were submitted
to a nested cross-validation scheme
for model selection and model assessment.
In the inner CV layer, the logistic regression estimators
have been trained in a one-versus-rest design that
distinguishes each class from
the respective 17 other classes
(number of maximal iterations=$100$, tolerance=$0.001$).
In the outer CV layer, grid search
selected among candidates for the respective $\lambda$ parameter
by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
Importantly, the thus selected sparse logistic regression classifier was
evaluated on an identical test set in all settings.

@@

\paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).
%
The stratified and shuffled training data were submitted
to a nested cross-validation scheme
for model selection and model assessment.
In the inner CV layer, the logistic regression estimators
have been trained in a one-versus-rest design that
distinguishes each class from
the respective 17 other classes
(number of maximal iterations=$100$, tolerance=$0.001$).
In the outer CV layer, grid search
selected among candidates for the respective $\lambda$ parameter
by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
Importantly, the thus selected sparse logistic regression classifier was
evaluated on an identical test set in all settings.

Three observations have been made.
In the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5,11\% as well as
better support recovery with weight maps already much closer
to the class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
%
Across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representatives among class-responsive predictor groups
due to the $\ell_1$-norm penalization.
%
In the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhance the similarity
to the weight maps even in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator reached this generalization performance
only with bigger input data quantities.

\begin{figure*}
\begin{centering}
\includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other.
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable predictive model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}

\paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\frac{1}{3}$ to $\eta_g$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

\begin{table*}[]
\centering
\caption{18-class out-of-sample performance by region-network emphasis}
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\textbf{Reg-Net Ratio} & 500 & 100  & 50   & 10   & 5    & 2    & 1    & $\frac{1}{2}$  & $\frac{1}{5}$  & $\frac{1}{10}$ & $\frac{1}{50}$ & $\frac{1}{100}$ & $\frac{1}{250}$ & $\frac{1}{500}$ & $\frac{1}{1000}$ \\ \hline
\textbf{Accuracy {[}\%{]}}      & 3.8          & 41.7 & 63.7 & 89.9 & 89.9 & 88.5 & 87.9 & 87.9 & 87.9 & 87.8 & 88.6 & 87.8  & 85.6  & 67.2  & 39,3  
\end{tabular}
\label{table_reg_net_ratio}
\end{table*}

@@

\paragraph{Acknowledgment.}
{\small The research leading to these results has received funding from the
European Union Seventh Framework Programme (FP7/2007-2013)
under grant agreement no. 604102 (Human Brain Project).
Data were provided by the Human Connectome Project.
Further support was received from
the German National Academic Foundation (D.B.),
the German Research Foundation (BZ2/2-1 and BZ2/3-1 to D.B.),
and the MetaMRI associated team (B.T., G.V.).
}

>>>>

\paragraph{Hyperparameter optimization.}
The stratified and shuffled training data were submitted
to a nested cross-validation scheme
for model selection and model assessment.
In the inner CV layer, the logistic regression estimators
have been trained in a one-versus-rest design that
distinguishes each class from
the respective 17 other classes
(number of maximal iterations=$100$, tolerance=$0.001$).
In the outer CV layer, grid search
selected among candidates for the respective $\lambda$ parameter
by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
Importantly, the thus selected sparse logistic regression classifier was
evaluated on an identical test set in all settings.

@@

\paragraph{Benchmarking hierarchical tree sparsity against
common sparsity penalties.}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation.
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with sparse group sparsity
imposes a structured $\ell_1/\ell_2$ block norm with additional $\ell_1$ term
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
% The stratified and shuffled training data were submitted
% to a nested cross-validation scheme
% for model selection and model assessment.
% In the inner CV layer, the logistic regression estimators
% have been trained in a one-versus-rest design that
% distinguishes each class from
% the respective 17 other classes
% (number of maximal iterations=$100$, tolerance=$0.001$).
% In the outer CV layer, grid search
% selected among candidates for the respective $\lambda$ parameter
% by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
% Importantly, the thus selected sparse logistic regression classifier was
% evaluated on an identical test set in all settings.

@@

\paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
were quantitatively compared (Figure \ref{fig_dataratio}).
Region-network priors should bias model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP task data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 neural activity maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).

% results: accuracy
Regarding classification performances on the always same test set,
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved
83.6\% versus 88.7\% (20\% of training data),
85.0\% versus 89.2\% (40\%),
86.8\% versus 89.8\% (60\%),
88.9\% versus 90.3\% (80\%),
88.6\% versus 89.7\% (100\%) accuracy.
% results: sparsity
Regarding the model sparsity,
the model weights $w$ of both penalized estimators were
was measured by $s = \frac{||w||_1}{||w||_F}$ for each of the 18 classes.
The $\ell_1$-penalized logistic regression
yielded the mean (+/- standard deviation) sparsity $s$ 
50.0 (+/- 2.6), 45.4 (+/- 2.3), 40.0 (+/- 2.4), 30.9 (+/- 2.0), and 24.0 (+/- 2.3)
after model fitting with 20\% to 100\% training data.
The hierarchical-tree-penalized logistic regression
yield the sparsity $s$ 
163.2 (+/- 0.7), 160.2 (+/- 1.8), 132.1 (+/- 3.3), 116.2 (+/- 4.4), and $88.4$ (+/- $8.4$)
after fitting 20\% to 100\% of the training data.
% results: support recovery
Finally, the support recovery was quantified by
Pearson correlation $\rho$ between vectors of
the zscored model coefficients
and
the zscored sample maps for each class.
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved a mean $\rho$ (+/- standard deviation) of
0.10 (+/- 0.03) versus 0.13 (+/- 0.04),
0.11 (+/- 0.03) versus 0.13 (+/- 0.05),
0.13 (+/- 0.04) versus 0.17 (+/- 0.05),
0.16 (+/- 0.04) versus 0.22 (+/- 0.05), and
0.19 (+/- 0.05) versus 0.29 (+/- 0.06)
from 20\% to 100\% training data.

Three observations have been made.
In the data-scarce scenario (i.e., 1/5 of available training data),
hierarchical tree sparsity achieved the biggest advantage
in out-of-sample performance by 5,11\% as well as
better support recovery with weight maps already much closer
to the class averages
\cite{varoquaux2012small}.
In the case of scarce training data, which is typical for the brain imaging domain,
regularization by region-network priors thus allowed for
more effective extraction of classification-relevant structure
from the neural activity maps.
%
Across training data fractions,
the weight maps from ordinary logistic regression exhibited
higher variance and more zero coefficients
than hierarchical tree logistic regression.
Given the usually high multicollinearity in neuroimaging data,
this observation is likely to reflect instable selection of
representatives among class-responsive predictor groups
due to the $\ell_1$-norm penalization.
%
In the data-rich scenario (i.e., entire training data used for model fitting),
neurobiologically informed logistic regression
profited more from the increased information quantities than
neurobiologically naive logistic regression.
That is, the region-network priors actually further enhance the similarity
to the weight maps even in abundant input data.
This was the case although
the maximal classification performance of $\approx$90\% has already
been reached with small training data fractions by the structured estimator.
In contrast, 
the unstructured estimator reached this generalization performance
only with bigger input data quantities.
%
\begin{figure*}
\begin{centering}
\includegraphics[width=0.90\textwidth]{../figures/reg_net_ratio_colin.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Support recovery as a function of
region and network emphasis.}
The relative impact of the region and network priors
on model selection
is systematically varied against each other.
This region-network ratio (\textit{upper fractions}) weighted voxel groups
to priviledge sparse models in function space
that acknowledge known brain region neighborhoods
(\textit{left columns}) or
known brain networks neighborhoods
(\textit{right columns}).
Among the 18 classes, the model weights are shown for the psychological
tasks (\textit{from top to bottom}): tongue movement, listening stories,
taking somebody else's perspective (ToM, "theory of mind"),
as well as
viewing locations and tools.
The 18-class out-of-sample accuracy is indicated
on the \textit{bottom} and
the class-wise mean neural activty
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
%
Different emphasis on regions versus networks
in hierarchical structured sparsity can
yield comparable model performance.
%
Favoring region versus network structure during model selection
recovers complementary aspects of the neural activity pattern.
%
Equal region and network emphasis yields more dispersed,
less interpretable predictive model choices.
}
\label{fig_regnetratio}
\end{centering}
\end{figure*}
%
\paragraph{Support recovery as a function of region-network emphasis.}
Finally, the relative importance of the
region and network priors within the hierarchical tree prior
was quantified (Figure \ref{fig_regnetratio}).
The group weight $\eta_g$ of region priors was multiplied with a
region-network ratio, while the
group weight $\eta_g$ of network priors was divided by that
region-network ratio. For instance, a region-network ratio of 3
increased the relative importance of known region structure
by multiplying $\frac{3}{1}$ to
$\eta_g$ of all region group penalties
and multiplying
$\frac{1}{3}$ to $\eta_g$ of all network group penalties
(Table \ref{table_reg_net_ratio}).

\begin{table*}[]
\centering
\caption{Out-of-sample performance by region-network emphasis}
\begin{tabular}{l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\textbf{Reg-Net Ratio} & 500 & 100  & 50   & 10   & 5    & 2    & 1    & $\frac{1}{2}$  & $\frac{1}{5}$  & $\frac{1}{10}$ & $\frac{1}{50}$ & $\frac{1}{100}$ & $\frac{1}{250}$ & $\frac{1}{500}$ & $\frac{1}{1000}$ \\ \hline
\textbf{Accuracy {[}\%{]}}      & 3.8          & 41.7 & 63.7 & 89.9 & 89.9 & 88.5 & 87.9 & 87.9 & 87.9 & 87.8 & 88.6 & 87.8  & 85.6  & 67.2  & 39,3  
\end{tabular}
\label{table_reg_net_ratio}
\end{table*}

@@

\paragraph{Acknowledgment.}
{\small Data were provided by the Human Connectome Project.
}

% {\small The research leading to these results has received funding from the
% European Union Seventh Framework Programme (FP7/2007-2013)
% under grant agreement no. 604102 (Human Brain Project).
% Data were provided by the Human Connectome Project.
% Further support was received from
% the German National Academic Foundation (D.B.),
% the German Research Foundation (BZ2/2-1 and BZ2/3-1 to D.B.),
% and the MetaMRI associated team (B.T., G.V.).
% }
